\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-16 paper.tex

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
Following the above definitions, corresponding data encoding for
this gARG is shown in Tab.~\ref{tab-gARG-data} (Appendix~\ref{sec-gARG-data}).

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
(A) An example ARG simulated from a diploid Wright-Fisher model.
(B) Remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass nodes in local trees in which they are unary.
In each case, the graph is shown on the left
and corresponding local trees on the right.
Graph nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent;
see the text for more details.
}
\end{figure}

@@

\clearpage
\section{Data encoding of the genome ARG}
\label{sec-gARG-data}

\begin{table}[ht]
    \caption{\label{tab-gARG-data}
    Data encoding of the genome ARG shown in Fig.~\ref{fig-arg-in-pedigree} with the
	node table on the left (having node and time columns) and the edge table on the
	right (having child, parent, and intervals columns).
    }
    \begin{center}
        \begin{tabular}{c|ccc|c|c}
			\multicolumn{2}{c}{Node table}& ~ &\multicolumn{3}{c}{Edge table}\\
			\cline{1-2} \cline{4-6}
            Node & Time & ~ & Child & Parent & Intervals\\
            \cline{1-2} \cline{4-6}
            $\noderef{a}$ & 0 & ~ & $\noderef{a}$ & $\noderef{e}$ & $\{[0,2)\}$\\
            $\noderef{b}$ & 0 & ~ & $\noderef{a}$ & $\noderef{f}$ & $\{[2,10)\}$\\
            $\noderef{c}$ & 0 & ~ & $\noderef{b}$ & $\noderef{g}$ & $\{[0,10)\}$\\
            $\noderef{d}$ & 0 & ~ & $\noderef{c}$ & $\noderef{f}$ & $\{[0,7)\}$\\
            $\noderef{e}$ & 1 & ~ & $\noderef{c}$ & $\noderef{e}$ & $\{[7,10)\}$\\
            $\noderef{f}$ & 1 & ~ & $\noderef{d}$ & $\noderef{h}$ & $\{[0,10)\}$\\
            $\noderef{g}$ & 1 & ~ & $\noderef{e}$ & $\noderef{i}$ & $\{[0,10)\}$\\
            $\noderef{i}$ & 1 & ~ & $\noderef{f}$ & $\noderef{k}$ & $\{[0,10)\}$\\
            $\noderef{k}$ & 2 & ~ & $\noderef{g}$ & $\noderef{i}$ & $\{[0,10)\}$\\
            $\noderef{n}$ & 3 & ~ & $\noderef{h}$ & $\noderef{k}$ & $\{[0,10)\}$\\
            \multicolumn{2}{c}{~}& ~ & $\noderef{i}$ & $\noderef{n}$ & $\{[0,10)\}$\\
            \multicolumn{2}{c}{~}& ~ & $\noderef{k}$ & $\noderef{n}$ & $\{[0,10)\}$
        \end{tabular}
    \end{center}
\end{table}

\clearpage
\section{TODO}
\label{sec-x}

% TODO: we don't seem to cite this figure at all!?

\begin{figure}[ht]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
	\end{center}
	\caption{\label{fig-simplification-with-edges}
	The four levels of ARG simplification from Fig.~\ref{fig-simplification}A--D, with edge intervals shown.
	}
\end{figure}

>>>>

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
Following the above definitions, corresponding data encoding for
this gARG is shown in Tab.~\ref{tab-gARG-data}.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
(A) An example ARG simulated from a diploid Wright-Fisher model.
(B) Remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass nodes in local trees in which they are unary.
In each case, the graph is shown on the left
and corresponding local trees on the right.
% Repetition here, but I'm thinking about the casual "I'm just reading
% the captions reader" who might not realise that there's an essential
% ingredient missing from the graph version
In the interest of visual clarity, inheritance intervals are not shown
on the graph edges; Supplementary Fig.~\ref{fig-simplification-with-edges}
shows the graphs with these inheritance intervals included.
Graph nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent;
see the text for more details.
}
\end{figure}

@@

\clearpage
\renewcommand\thefigure{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand\thetable{S\arabic{table}}
\setcounter{table}{0}

\section*{Supplementary Material}

\begin{table}[ht]
    \begin{center}
        \begin{tabular}{c|ccc|c|c}
			\multicolumn{2}{c}{Node table}& ~ &\multicolumn{3}{c}{Edge table}\\
			\cline{1-2} \cline{4-6}
            Node & Time & ~ & Child & Parent & Intervals\\
            \cline{1-2} \cline{4-6}
            $\noderef{a}$ & 0 & ~ & $\noderef{a}$ & $\noderef{e}$ & $\{[0,2)\}$\\
            $\noderef{b}$ & 0 & ~ & $\noderef{a}$ & $\noderef{f}$ & $\{[2,10)\}$\\
            $\noderef{c}$ & 0 & ~ & $\noderef{b}$ & $\noderef{g}$ & $\{[0,10)\}$\\
            $\noderef{d}$ & 0 & ~ & $\noderef{c}$ & $\noderef{f}$ & $\{[0,7)\}$\\
            $\noderef{e}$ & 1 & ~ & $\noderef{c}$ & $\noderef{e}$ & $\{[7,10)\}$\\
            $\noderef{f}$ & 1 & ~ & $\noderef{d}$ & $\noderef{h}$ & $\{[0,10)\}$\\
            $\noderef{g}$ & 1 & ~ & $\noderef{e}$ & $\noderef{i}$ & $\{[0,10)\}$\\
            $\noderef{i}$ & 1 & ~ & $\noderef{f}$ & $\noderef{k}$ & $\{[0,10)\}$\\
            $\noderef{k}$ & 2 & ~ & $\noderef{g}$ & $\noderef{i}$ & $\{[0,10)\}$\\
            $\noderef{n}$ & 3 & ~ & $\noderef{h}$ & $\noderef{k}$ & $\{[0,10)\}$\\
            \multicolumn{2}{c}{~}& ~ & $\noderef{i}$ & $\noderef{n}$ & $\{[0,10)\}$\\
            \multicolumn{2}{c}{~}& ~ & $\noderef{k}$ & $\noderef{n}$ & $\{[0,10)\}$
        \end{tabular}
    \end{center}
    % NOTE: I know  the "correct" placement for table captions is on top, but I
    % always find this quite weird (since many of our figs are quite "table
    % like" anyway. If anyone feels strongly about putting the caption above
    % the table please just move it back.
    \caption{\label{tab-gARG-data}
    Data encoding of the genome ARG shown in Fig.~\ref{fig-arg-in-pedigree} with the
	node table on the left (having node and time columns) and the edge table on the
	right (having child, parent, and intervals columns).
    }
\end{table}

\begin{figure}[]
	\begin{center}
		\includegraphics[width=0.8\textwidth]{illustrations/simplification-with-edges.pdf}
	\end{center}
	\caption{\label{fig-simplification-with-edges}
	Example ARGs Fig.~\ref{fig-simplification}A--D,
    with edges annotated with inheritance intervals.
	}
\end{figure}
